\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-04-04 additional_files/additional_file_1.tex
Invert Methods and Results and change figure numbers accordingly
Change the word Additional to Supplemental to be concordant with The Plant Genome's standard
Format all parts of the manuscript and supplemental material (except references) as required by The Plant Genome

\documentclass[12pt]{report}

@@

@@

\chapter*{Supplemental data to ``\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean''}
\label{supdata}

@@

% List of authors
\textbf{Authors}: Marc-André Lemay, Maxime de Ronne, Richard Bélanger, François Belzile

@@

\subsection*{Comparison of WGS and SoySNP50K genotypes}
\label{annexe-sv-gwas-soysnp50k}
When we first analyzed the results from the GWAS analysis on all 389 samples,
we found discrepancies between the haplotypes at some loci and the expected
phenotype that should correspond to that haplotype for simple traits such as
flower color and seed coat color. Upon further analysis, we found that a
mismatch between the sequencing data and the identity of the accession was the
most probable cause of these discrepancies. In order to identify the samples
for which this could be the case, we compared the genotypes of SNPs obtained
from the SoySNP50K array \citep{song2013} to those predicted from the Illumina WGS
data. We did this for the 385 samples that had a match in the GRIN database. Briefly, we
retrieved SoySNP50K genotype calls from SoyBase \citep{grant2010} and extracted
all SNPs with a MAF $\geq$ 0.1 in our population. We then translated the positions
of these SNPs from assembly version 2 to assembly version 4 of Williams by
finding exact matches of 41-bp sequences surrounding the SNP positions. The
32,852 SNPs for which unambiguous positions could be found on genome assembly
version 4 were genotyped from the mapped WGS reads using \texttt{bcftools
mpileup} and \texttt{bcftools call}. We then used \texttt{bcftools gtcheck} to
compute the concordance between the WGS genotype calls and the
SoySNP50K genotype dataset. We identified 24 samples
with < 90\% concordance between the two genotype datasets and removed them from
the dataset. We also excluded an additional sample for which concordance was over 90\%
but poor raw sequencing data quality (\%GC content) suggested contamination.
The concordance between WGS data and SoySNP50K genotypes is included in
Additional file 3 and shown in Figure \ref{concordance-histogram}.

@@

% Figures mentioned in the main text
% Figures for flower color
\manhattanplot{flower_color}{flower color}{\getvar{flower_color_nsamples}}
\signalplot{flower_color_W1}{flower color}{\emph{W1}}%
{Vertical dotted lines indicate the location of the \getvar{flower_color_W1_gene} gene associated with the locus.}
\geneplot{flower_color_W1}{flower color}{\emph{W1}}{\getvar{flower_color_W1_gene} gene}%
{Vertical dotted lines in panels (c) and (d) indicate the location of the causal SV at this locus.}
\ldplot{flower_color}{flower color}{\getvar{flower_color_ldkmers}}
%\kmerplot{flower_color_W1}

@@

% Histogram with concordance rates
\begin{cfigure}
	\includegraphics[width = 6in]{concordance_histogram}
	\caption{Histogram of concordance rates between genotypes derived from WGS data and SoySNP50K genotypes for 385 soybean accessions.
	The vertical line indicates the 0.9 concordance threshold that was used for filtering out mismatching samples.}
	\label{concordance-histogram}

\end{cfigure}

>>>>

\documentclass[12pt]{article}

@@

\title{Supplementary material to ``\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean''}
\date{}

\author{Marc-André Lemay \and
Maxime de Ronne \and
Richard Bélanger \and
François Belzile}

@@

\maketitle

@@

@@

\subsection*{Comparison of WGS and SoySNP50K genotypes}
\label{annexe-sv-gwas-soysnp50k}
When we first analyzed the results from the GWAS analysis on all 389 samples,
we found discrepancies between the haplotypes at some loci and the expected
phenotype that should correspond to that haplotype for simple traits such as
flower color and seed coat color. Upon further analysis, we found that a
mismatch between the sequencing data and the identity of the accession was the
most probable cause of these discrepancies. In order to identify the samples
for which this could be the case, we compared the genotypes of SNPs obtained
from the SoySNP50K array \citep{song2013} to those predicted from the Illumina WGS
data. We did this for the 385 samples that had a match in the GRIN database. Briefly, we
retrieved SoySNP50K genotype calls from SoyBase \citep{grant2010} and extracted
all SNPs with a MAF $\geq$ 0.1 in our population. We then translated the positions
of these SNPs from assembly version 2 to assembly version 4 of Williams by
finding exact matches of 41-bp sequences surrounding the SNP positions. The
32,852 SNPs for which unambiguous positions could be found on genome assembly
version 4 were genotyped from the mapped WGS reads using \texttt{bcftools
mpileup} and \texttt{bcftools call}. We then used \texttt{bcftools gtcheck} to
compute the concordance between the WGS genotype calls and the
SoySNP50K genotype dataset. We identified 24 samples
with < 90\% concordance between the two genotype datasets and removed them from
the dataset. We also excluded an additional sample for which concordance was over 90\%
but poor raw sequencing data quality (\%GC content) suggested contamination.
The concordance between WGS data and SoySNP50K genotypes is included in
Supplemental file 3 and shown in Figure \ref{concordance-histogram}.

@@

% Figures mentioned in the main text

% Histogram with concordance rates
\begin{cfigure}
	\includegraphics[width = 6in]{concordance_histogram}
	\caption{Histogram of concordance rates between genotypes derived from WGS data and SoySNP50K genotypes for 385 soybean accessions.
	The vertical line indicates the 0.9 concordance threshold that was used for filtering out mismatching samples.}
	\label{concordance-histogram}
\end{cfigure}

% Figures for flower color
\manhattanplot{flower_color}{flower color}{\getvar{flower_color_nsamples}}
\signalplot{flower_color_W1}{flower color}{\emph{W1}}%
{Vertical dotted lines indicate the location of the \getvar{flower_color_W1_gene} gene associated with the locus.}
\geneplot{flower_color_W1}{flower color}{\emph{W1}}{\getvar{flower_color_W1_gene} gene}%
{Vertical dotted lines in panels (c) and (d) indicate the location of the causal SV at this locus.}
\ldplot{flower_color}{flower color}{\getvar{flower_color_ldkmers}}
%\kmerplot{flower_color_W1}

@@

### 2023-04-04 additional_files/manuscript.tex
Format all parts of the manuscript and supplemental material (except references) as required by The Plant Genome

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support

@@

\title{k-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean}
\date{}

\author{Marc-André Lemay\textsuperscript{1,2,3} \and
Maxime de Ronne\textsuperscript{1,2,3} \and
Richard Bélanger\textsuperscript{1,2,3} \and
François Belzile\textsuperscript{1,2,3}}

@@

@@

Genome-wide association studies (GWAS) are powerful statistical methods that
detect associations between genotype and phenotype at genome scale. Despite
their power, GWAS frequently fail to pinpoint the causal variant or the gene
controlling a given trait in crop species. Assessing genetic
variants beyond single-nucleotide polymorphisms (SNPs) could alleviate this
problem, for example by including structural variants (SVs).

In this study, we tested the potential of SV- and \emph{k}-mer-based GWAS in
soybean by applying these methods to 13 traits.  We also performed
conventional GWAS analysis based on SNPs and small indels.  We assessed the performance of each GWAS approach based on
results at loci for which the causal genes or variants were known from
previous genetic studies. We found that \emph{k}-mer-based GWAS was the
most versatile approach and the best at pinpointing causal variants or
candidate genes based on the most significantly associated
\emph{k}-mers.  Moreover, \textit{k}-mer-based analyses identified
promising candidate genes for loci related to pod color, pubescence
form, and resistance to \textit{Phytophthora sojae}. In
our dataset, SV-based GWAS did not add value compared to
\textit{k}-mer-based GWAS and may not be worth the time and
computational resources required to genotype SVs at population scale.

Despite promising results, significant challenges remain regarding the
downstream analysis of \emph{k}-mer-based GWAS.  Notably, better
methods are needed to associate significant \emph{k}-mers with sequence
variation. Together, our results suggest that coupling \emph{k}-mer-
and SNP/indel-based GWAS is a powerful approach for discovering
candidate genes in crop species.

\section*{Keywords}
Genome-wide association studies
\textit{k}-mer GWAS
Soybean genomics
Crop genomics
\textit{Phytophthora sojae}
Structural variation

@@

@@

\section*{Supplemental Files}

\subsection*{Supplemental file 1 --- Supplemental file 1.pdf}
Supplemental methods and results,
including Supplemental Tables S1 to S11
and Supplemental Figures S1 to S60 (PDF 12.9 MB)

\subsection*{Supplemental file 2 --- Supplemental file 2.csv}
Metadata on the sequencing runs retrieved
from the Sequence Read Archive (SRA) (CSV 71 KB)

\subsection*{Supplemental file 3 --- Supplemental file 3.csv}
Metadata on individual accessions following mapping
and comparison to SoySNP50K data (CSV 36KB)

\subsection*{Supplemental file 4 --- Supplemental file 4.csv}
Phenotypic dataset used for GWAS analyses (CSV 135 KB)

\section*{Acknowledgements}%% if any

@@

\section*{Abbreviations}%% if any
BAM: Binary alignment map format;
CDW: corrected dry weight;
CHS: chalcone synthase;
CNV: copy number variant, copy number variation;
GRIN: Germplasm Resources Information Network;
GWAS: Genome-wide association study;
HPS: Hydrophobic protein from soybean;
LD: Linkage disequilibrium;
MAF: Minor allele frequency;
MATE: Multidrug and toxic compound extrusion;
MLM: Mixed linear model;
NCBI: National Center for Biotechnology Information;
SNP: Single-nucleotide polymorphism;
SRA: Sequence Read Archive;
SV: Structural variant, structural variation;
VCF: Variant call format;
WGS: Whole-genome (re)sequencing

@@

\section*{Ethics approval and consent to participate}%% if any
Not applicable.

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Consent for publication}%% if any
Not applicable.

\section*{Authors' contributions}
Conception and design of the study: MAL, MDR, RB, FB.
Phenotypic data acquisition (resistance to \textit{Phytophthora sojae}): MDR, RB.
Data analysis: MAL.
Data interpretation: MAL, FB.
Software writing: MAL.
Manuscript drafting: MAL, FB.
All authors have revised the manuscript and approved its submission.

@@

% Table 1 with results on all loci with cloned genes
\section*{Tables}
\begin{landscape}
\begin{table}
	\caption[Performance of three GWAS approaches at detecting causal genes/variants]%
	{Performance of three GWAS approaches at detecting causal genes or variants at cloned loci.}
	\label{loci-table}

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc} %unicode support

@@

\title{\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean}
\date{}

\author{Marc-André Lemay\textsuperscript{1,2,3} \and
Maxime de Ronne\textsuperscript{1,2,3} \and
Richard Bélanger\textsuperscript{1,2,3} \and
François Belzile\textsuperscript{1,2,3,*}}

@@

\textsuperscript{*} Corresponding author: francois.belzile@fsaa.ulaval.ca

\clearpage

\section*{Core ideas}

\begin{itemize}
	\item We compared SNP-, SV- and \textit{k}-mer-based GWAS approaches
		on 13 traits using a population of 363 cultivated soybean lines.
	\item \textit{k}-mer-based GWAS performed better than the other two approaches
		at pinpointing causal variants and genes.
	\item We developed computational tools to facilitate the downstream analysis
		of \textit{k}-mer-based GWAS.
	\item Performing SV-based GWAS in soybean may not be worth the investment
		of time and computational resources.
\end{itemize}

\section*{Abbreviations}
BAM, Binary alignment map format;
CDW, corrected dry weight;
CHS, chalcone synthase;
CNV, copy number variant, copy number variation;
GRIN, Germplasm Resources Information Network;
GWAS, Genome-wide association study;
HPS, Hydrophobic protein from soybean;
LD, Linkage disequilibrium;
MAF, Minor allele frequency;
MATE, Multidrug and toxic compound extrusion;
MLM, Mixed linear model;
NCBI, National Center for Biotechnology Information;
SNP, Single-nucleotide polymorphism;
SRA, Sequence Read Archive;
SV, Structural variant, structural variation;
VCF, Variant call format;
WGS, Whole-genome (re)sequencing

@@

Genome-wide association studies (GWAS) are powerful statistical methods that
detect associations between genotype and phenotype at genome scale. Despite
their power, GWAS frequently fail to pinpoint the causal variant or the gene
controlling a given trait in crop species. Assessing genetic
variants beyond single-nucleotide polymorphisms (SNPs) could alleviate this
problem, for example by including structural variants (SVs).
In this study, we tested the potential of SV- and \emph{k}-mer-based GWAS in
soybean by applying these methods to 13 traits.  We also performed
conventional GWAS analysis based on SNPs and small indels.  We assessed the performance of each GWAS approach based on
results at loci for which the causal genes or variants were known from
previous genetic studies. We found that \emph{k}-mer-based GWAS was the
most versatile approach and the best at pinpointing causal variants or
candidate genes based on the most significantly associated
\emph{k}-mers.  Moreover, \textit{k}-mer-based analyses identified
promising candidate genes for loci related to pod color, pubescence
form, and resistance to \textit{Phytophthora sojae}. In
our dataset, SV-based GWAS did not add value compared to
\textit{k}-mer-based GWAS and may not be worth the time and
computational resources required to genotype SVs at population scale.
Despite promising results, significant challenges remain regarding the
downstream analysis of \emph{k}-mer-based GWAS.  Notably, better
methods are needed to associate significant \emph{k}-mers with sequence
variation. Together, our results suggest that coupling \emph{k}-mer-
and SNP/indel-based GWAS is a powerful approach for discovering
candidate genes in crop species.

\clearpage

@@

\clearpage

@@

\clearpage

\section*{Acknowledgments}

@@

\section*{Conflict of interest}
The authors declare no conflict of interest.

\section*{ORCID}

\begin{description}
	\item[Marc-André Lemay] \url{https://orcid.org/0000-0003-2484-8447}
	\item[Maxime de Ronne] \url{https://orcid.org/0000-0003-4062-3512}
	\item[Richard Bélanger] \url{https://orcid.org/0000-0001-8549-975X}
	\item[François Belzile] \url{https://orcid.org/0000-0001-6893-4529}
\end{description}

\section*{Supplemental material}

\begin{description}
	\item[Supplemental file 1.pdf] Supplemental methods and results, including
		Supplemental Tables S1 to S11 and Supplemental Figures S1 to
		S60.

	\item[Supplemental file 2.csv] Metadata on the sequencing runs retrieved
		from the Sequence Read Archive (SRA).

	\item[Supplemental file 3.csv] Metadata on individual accessions following
		mapping and comparison to SoySNP50K data.

	\item[Supplemental file 4.csv] Phenotypic dataset used for GWAS analyses.
\end{description}

@@

\section*{Author contributions}
Conceptualization: MAL, MDR, RB, FB.
Data curation: MAL, MDR.
Methodology: MAL, FB.
Investigation: MAL, MDR.
Formal analysis: MAL.
Funding acquisition: RB, FB.
Data interpretation: MAL, FB.
Software: MAL.
Supervision: RB, FB.
Visualization: MAL.
Writing -- original draft: MAL, FB.
Writing -- review \& editing: MAL, MDR, RB, FB.

\clearpage

@@

\clearpage

\section*{Tables}

% Table 1 with results on all loci with cloned genes
\begin{landscape}
\begin{table}
	\caption[Performance of three GWAS approaches at detecting causal genes/variants]%
	{Performance of three GWAS approaches at detecting causal genes or variants at cloned loci.}
	\label{loci-table}
